---
title: "Class 8: Breast Cancer Mini Project"
author: "Lana (PID: A17013518)"
format: pdf
---

Before we get stuck into project work we will have a quick look at applying PCA to some example RNASeq data (tail end of lab 7).

Read the data:

```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)
```

> Q. How many genes are in this data?

```{r}
nrow(rna.data)
```

## Run PCA

```{r}
## Again we have to take the transpose of our data 
pca <- prcomp(t(rna.data), scale=TRUE)
 
## Simple un polished plot of pc1 and pc2
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")
```

```{r}
summary(pca)
```

```{r}
pca$x
```

```{r}
# We have 5 wt and 5 ko samples
mycols <- c(rep("pink", 5), rep("yellow", 5))
mycols

plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", col = mycols)
```

I could examine which genes contribute most to this first PC

```{r}
head(sort(abs(pca$rotation[,1]), decreasing = T))
```

# Analysis of Breast Cancer Data

First we will read the data:

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names=1)
head(wisc.df)
```

Note that the first column here wisc.df$diagnosis is a pathologist provided expert diagnosis. 

```{r}
diagnosis <- as.factor(wisc.df$diagnosis)
diagnosis
```

To make sure we don’t accidentally include this in our analysis, lets create a new data.frame that omits this first column

```{r}
# We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]
head(wisc.data)
```

> Q1. How many observations are in this dataset?

```{r}
ncol(wisc.data)
```


> Q2. How many of the observations have a malignant diagnosis?

```{r}
table(wisc.df$diagnosis)
```


> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
length(grep("_mean", colnames(wisc.data)))
```

## Principal Component Analysis

Here we will use `prcomp()` on the `wisc.data` object - the one without diagonisis column.

First, we have to decide wether to use the `scale = T` argument when we run `prcomp()`

We can look at the means and sd of each column. If they are similar then we are all good to go. If we should use `scale = T` 

```{r}
colMeans(wisc.data)
```

```{r}
apply(wisc.data, 2, sd)
```

These are very different so we should scale = T

```{r}
wisc.pr <- prcomp(wisc.data, scale = T)
summary(wisc.pr)
```

> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

44.27%

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3 PCs capture 72.6%

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

7 PCs capture 91.0%

### Plotting the PCA results

```{r}
# biplot(wisc.pr)
```

```{r}
attributes(wisc.pr)
```

```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2], col = diagnosis)
```

```{r}
library(ggplot2)

pc <- as.data.frame(wisc.pr$x)

ggplot(pc) +
  aes(PC1, PC2, col = diagnosis) +
  geom_point()
```

## Communicating PCA Results

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

```{r}
wisc.pr$rotation["concave.points_mean", 1]
```

> Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

```{r}
tbl <- summary(wisc.pr)
which(tbl$importance[3,] > 0.8) [1]
```

## Hierarchical clustering

The main function for hierarical clustering is `hclust()` it takes a distance matrix as input.

```{r}
d <- dist(scale(wisc.data))
wisc.hclust <- hclust(d)
plot(wisc.hclust)
```

```{r}
plot(wisc.hclust)
abline(h=18, col = "pink")
grps <- cutree(wisc.hclust, h=18)
table(grps)
```

Come back here later to see how our cluster grps respond to M or B groups. 

```{r}
ggplot(pc) +
  aes(PC1, PC2, col = diagnosis) +
  geom_point()
```

# Clustering on PCA Results

```{r}
d.pc <- dist(wisc.pr$x[,1:3])

wisc.pr.hclust <- hclust( d.pc, method = "ward.D2")
plot(wisc.pr.hclust)
abline(h = 80, col = "pink")
```

```{r}
grps <- cutree(wisc.pr.hclust, h = 80)
table(grps)
```

We can use `table()` to make a cross-table and count table

```{r}
table(diagnosis)
```

```{r}
table(grps, diagnosis)
```

The results indicate that our cluster 1 mostly captures cancer (M) and our cluster 2 mostly captures cancer (B)

## 7. Precidtion

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

plot

```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

> Q 18. 

Patient 2
