---
title: "Class05: Data Vis with ggplot"
author: "Lana (PID:A17013518)"
format: pdf
---

# Graphics systems in R

There are many graphics systems in R for making plots and figures.

We have already played a little with **base R** graphics and the `plot()` function.

Today we will start learning about a popular graphics package called `ggplot2()`.

This is an add on package - i.e. we need to install it. I install it (like i install any package) with the `install.packages()` function.


```{r}
plot(cars)
```


Before I can use the functions from a package I have to load up the package from "my library". We use the `library(ggplot2)` command to load it up.

```{r}
library(ggplot2)
ggplot(cars)
```

Every ggplot is made up of at least 3 things:
- data (the numbers  etc. that will go into your plot)
- aes  (how the columns of data map to the plot aesthetics)
- geoms (how the plot actually looks, points, lines, bars, etc.)


```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point()
```

For simple plots ggplot is more verbose - it takes more code - than base R plot.

Add some more layers to our ggplot:

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method="lm", color = "pink", se = F) +
  labs(title = "Stopping distance of old cars", subtitle = "A silly example plot") +
  theme_bw()
```

        
```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

```{r}
nrow(genes)
colnames(genes)
table(genes$State)
table(genes$State) / 5196
```

```{r}
p <- ggplot(genes) + 
    aes(x=Condition1, y=Condition2, col=State) +
    geom_point()
p
```

```{r}
p + labs(title="Gene Expression Changes Upon Drug Treatment") + scale_colour_manual(values=c("pink","yellow","purple"))
```

```{r}
# install.packages("dplyr")  ## un-comment to install if needed
library(gapminder)
library(dplyr)

gapminder_2007 <- gapminder %>% filter(year==2007)
```

```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp, color=continent, size=pop) + 
  geom_point(alpha=0.5)
```

```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp, color=pop) + 
  geom_point(alpha=0.8)
```

```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp, size=pop) + 
  geom_point(alpha=0.5)
```

```{r}
ggplot(gapminder_2007) +
  geom_point(aes(x=gdpPercap, y=lifeExp, size=pop), alpha=0.5) + 
  scale_size_area(max_size=10)
```

