---
title: "Class 18: Mini Project: Investigating Pertussis Resurgence"
author: "Lana (PID:A17013518)"
format: pdf
---

First we will examine and explore Pertussis case numbers in teh US as tracked by the CDC. 

We can use the datapasta package tos cape this data from the website into R:

```{r}
cdc <- data.frame(
  
  year = c(
    1922L,
    1923L,
    1924L,
    1925L,
    1926L,
    1927L,
    1928L,
    1929L,
    1930L,
    1931L,
    1932L,
    1933L,
    1934L,
    1935L,
    1936L,
    1937L,
    1938L,
    1939L,
    1940L,
    1941L,
    1942L,
    1943L,
    1944L,
    1945L,
    1946L,
    1947L,
    1948L,
    1949L,
    1950L,
    1951L,
    1952L,
    1953L,
    1954L,
    1955L,
    1956L,
    1957L,
    1958L,
    1959L,
    1960L,
    1961L,
    1962L,
    1963L,
    1964L,
    1965L,
    1966L,
    1967L,
    1968L,
    1969L,
    1970L,
    1971L,
    1972L,
    1973L,
    1974L,
    1975L,
    1976L,
    1977L,
    1978L,
    1979L,
    1980L,
    1981L,
    1982L,
    1983L,
    1984L,
    1985L,
    1986L,
    1987L,
    1988L,
    1989L,
    1990L,
    1991L,
    1992L,
    1993L,
    1994L,
    1995L,
    1996L,
    1997L,
    1998L,
    1999L,
    2000L,
    2001L,
    2002L,
    2003L,
    2004L,
    2005L,
    2006L,
    2007L,
    2008L,
    2009L,
    2010L,
    2011L,
    2012L,
    2013L,
    2014L,
    2015L,
    2016L,
    2017L,
    2018L,
    2019L,
    2020L,
    2021L
  ),
  cases = c(
    107473,
    164191,
    165418,
    152003,
    202210,
    181411,
    161799,
    197371,
    166914,
    172559,
    215343,
    179135,
    265269,
    180518,
    147237,
    214652,
    227319,
    103188,
    183866,
    222202,
    191383,
    191890,
    109873,
    133792,
    109860,
    156517,
    74715,
    69479,
    120718,
    68687,
    45030,
    37129,
    60886,
    62786,
    31732,
    28295,
    32148,
    40005,
    14809,
    11468,
    17749,
    17135,
    13005,
    6799,
    7717,
    9718,
    4810,
    3285,
    4249,
    3036,
    3287,
    1759,
    2402,
    1738,
    1010,
    2177,
    2063,
    1623,
    1730,
    1248,
    1895,
    2463,
    2276,
    3589,
    4195,
    2823,
    3450,
    4157,
    4570,
    2719,
    4083,
    6586,
    4617,
    5137,
    7796,
    6564,
    7405,
    7298,
    7867,
    7580,
    9771,
    11647,
    25827,
    25616,
    15632,
    10454,
    13278,
    16858,
    27550,
    18719,
    48277,
    28639,
    32971,
    20762,
    17972,
    18975,
    15609,
    18617,
    6124,
    2116
  )
                 )
head(cdc)
```

I want a plot of cases per year with ggplot.

```{r}
library(ggplot2)
```
 > Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.
 > Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc) +
  aes(year, cases) +
  geom_line() +
  geom_vline(xintercept = 1947, col = "pink") +
  geom_vline(xintercept = 1992, col = "violet") +
  geom_vline(xintercept = 2020, col = "turquoise")
```

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the aP vaccine it is very apparent that pertussis cases were increased. A couple possible explanations for this trend include: more sensitive PCR-based testing, vaccination hesitancy, bacterial evolution, and immunity in infants

## Access data from the CMI-PB project

This database (like many modern projects) uses an API to return JSON format data.

We will use the R package `jsonlite`.
```{r}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/subject", 
          simplifyVector = T)

head(subject)
```

> Q. How many wP  (the older whole-cell vaccine) indivisuals and aP (newer acellular vaccine) indivisuals are in this dataset?

```{r}
sum(subject$infancy_vac == "wP")
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

```{r}
subject$year_of_birth
```

## Side-note: Working with Dates

```{r}
library(lubridate)
```

```{r}
today()
```

```{r}
today() - ymd("2001-01-13")
```

```{r}
time_length( today() - ymd("2001-01-13"),  "years")
```
 
 > Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?
 
```{r}
subject$age <- time_length( today() - ymd(subject$year_of_birth), "years")
```

```{r}
library(dplyr)

ap <- subject %>% filter(infancy_vac == "aP")

round( summary( time_length( ap$age, "years" ) ) )
```

```{r}
wp <- subject %>% filter(infancy_vac == "wP")
round( summary( time_length( wp$age, "years" ) ) )
```

> Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```

> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=F) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in Years")
```

## Get more data from CMI-PB

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = T)
head(specimen)
```


```{r}
titer <- read_json("https://www.cmi-pb.org/api/v4/plasma_ab_titer", simplifyVector = T)
head(titer)
```

> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
library(dplyr)

meta <- inner_join(specimen, subject)
dim(meta)
head(meta)
```

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
```

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
 table(abdata$isotype)
```
 
 > Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?
 
```{r}
table(abdata$dataset)
```
 

## Our first exploratory plot

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() +
  xlim(0,75) +
  facet_wrap(vars(visit), nrow = 2)
```

> Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

The CMI-PB "Terminology Browser" was not working for me.

Can you facet or even just color by infancy_vac?

```{r}
ggplot(abdata) + 
  aes(MFI, antigen, col=infancy_vac) +
  geom_boxplot()
```

> Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
filter(igg, antigen=="OVA") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = F) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

```{r}
filter(igg, antigen=="FIM2/3") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = F) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

> Q16. What do you notice about these two antigens time courses and the PT data in particular?

I notice that PT levels rise over time and exceed the OVA. It also tends to peak around 5 adn then goes down after that. 

> Q17. Do you see any clear difference in aP vs. wP responses?

Same throughout both aP and wP.

```{r}
table(abdata$dataset)
```

Let's focus in on one dataset.

```{r}
abdata.21 <- filter(abdata, dataset == "2021_dataset")
table(abdata.21$dataset)
```

Focus on PT antigen

```{r}
pt.21 <- filter(abdata.21, isotype == "IgG", antigen == "PT")
```

Plot of 

```{r}
ggplot(pt.21) +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

